---
title: "Avicenna ESM Compliance Monitoring Tool"
author: "Robin Achterhof"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# About this ESM Compliance Monitoring Tool 

This document is a beginner-friendly template for monitoring Experience Sampling Method (ESM) compliance in studies that use **Avicenna** to collect data. It allows you to:

 - **track compliance** while your ESM study is running
 - include **one or multiple** different Avicenna questionnaires
 - perform **automatic checks** on things that may be going wrong
 - do all this **dynamically**, getting you quick updates while additional data is coming in
 - test everything using a built-in **demo dataset**

When you knit this document in R Markdown (`Ctrl+Shift+K`), it will give you a nice overview of common issues, compliance rates and visualizations. And if you want, the script can also generate an Excel file with participant‑level compliance (and optional payment information).

Of course, it is very well possible that there are some mistakes in this template. If you encounter any, please do let me know at [achterhof@essb.eur.nl](mailto:achterhof@essb.eur.nl).

**Acknowledgements:** This entire tool is inspired by the ESM compliance management tools developed by Anne Buelow, developed for the [100 Days Of My Life](https://osf.io/a9g7c/overview) study, and the [PARADOX](https://osf.io/mwpt7/files/m2qwh) study. These tools were intended to use only for those studies, and have not been published.
In addition, generative AI has been used to help draft and hone the code throughout this script.


**To use this template:**

1. Place this .Rmd file in a folder of your choice
2. Create a subfolder named `/data`, and place your Avicenna `.csv` file(s) in this folder. The script will automatically detect and load the newest file, and archive any older files into `/old` 
3. Always update the study settings in the first code chunk below. 
4. Enable optional modules (ID matching, extra plots, Excel export) in that same first code chunk.
5. Knit the document.  

Note: This template assumes **standard Avicenna variable names**, including:

- `Participant.ID`  
- `Session.Scheduled.Time`  
- `Prompt.Time`  
- `Record.Time`  
- `Status`  

If your file uses different names, you must rename them before using this template.

If you have any questions about Avicenna data, please also take a look at the help available on https://forum.avicennaresearch.com/


# 1. User Inputs

This first section contains the **only settings you need to modify**: Information about days and number of beeps, a demo dataset to test this script with, and some toggles that allow you to use multiple Avicenna files, merge this file with an external file of participant ID's, add extra visualizations, and create an Excel export with relevant information.

## Study settings (required)

These settings can be adapted during the study, reflecting the current moment in the study^[It may be that you are in the middle of a current response window. It is recommended that you always refer to the last fully elapsed beep, to not underestimate compliance too much]. When the study has ended completely, you can just enter the last day of the study, and the last prompt of the day.


```{r study-settings}
# Folder where data is stored
data_path <- "data"

# Core study settings
# Note: if use_demo_dataset = TRUE, the demo is generated using these values.
# Adjust them to reflect the study you want to simulate.
study_day       <- 5
current_prompt  <- 3
beeps_per_day   <- 10
start_date      <- as.Date("2026-02-28")

```

## Demo mode (quick start)

If you want to try this template without your own data, enable the demo dataset here. If you want, you can also adjust the number of participants, days, and beeps per day in these simulated data.

```{r use-demo}
use_demo_dataset <- TRUE
# Demo dataset settings (only used when use_demo_dataset = TRUE)
demo_n_participants <- 30
demo_n_days        <- study_day        # mirrors your study settings
demo_beeps_per_day <- beeps_per_day    # mirrors your study settings
```

> **Using the demo?** Make sure the study settings above (`study_day`, 
> `beeps_per_day`, `start_date`) reflect the study you want to simulate — 
> the demo dataset is generated based on those values. The defaults are a 
> good starting point. Set `use_demo_dataset <- FALSE` when you are ready 
> to use your own data.

## Optional Modules (simple toggles)

**About the toggles:** These settings work like on/off switches, controlling whether optional modules in the template are executed.

Set a toggle to TRUE to activate a module.

If you are unsure, leave everything on FALSE.

```{r toggles}
use_multiple_files <- FALSE   # Merge multiple Avicenna Activities
use_excel_export   <- FALSE    # Export compliance to Excel
use_external_ids   <- FALSE    # Merge external participant IDs (requires external_ids.xlsx)
use_remuneration   <- FALSE    # Add payment calculations to Excel
```

## Advanced Settings (only relevant if toggles above are TRUE)

### Multiple Avicenna files

Only needed if `use_multiple_files = TRUE`.

```{r set-multiple-files}
activity_ids <- c(27337, 28137, 28138)   # e.g., c(27337, 28137, 28138)
```

**About the multiple files:** You only need to use this setting if you want to merge different data sets that you export from Avicenna (otherwise, you can just leave it blank). Each Avicenna questionnaire ("Activity") has a unique 5‑digit ID.If you have multiple questionnaires (e.g., morning, evening, weekend-only, weekly), simply add all their Activity IDs to activity_ids below.

You can find this ID in the file name of your export, e.g.: `study-4882-export-166-survey-responses-27337-2026-03-16-12-17-13.csv`. 
Here, **27337** would be the Activity ID. This number corresponds with the number you can find next to each separate questionnaire in the Avicenna dashboard. 

### External IDs

Only needed if `use_external_ids = TRUE`.

**About the external ID's:**
If you want to link Avicenna participant IDs to your own participant codes, set `use_external_ids = TRUE` and create a file named `external_ids.xlsx` with the columns `Participant.ID` (representing the Avicenna ID) and `External.ID` (representing the participant ID you want to link). Place this file in the same folder as this .Rmd file.

### Remuneration 

Only needed if `use_remuneration = TRUE`.

**About remuneration**: Some studies pay participants **per completed beep** (e.g., €0.50 per completed prompt), or a different amount when a certain **threshold** is reached (e.g., €20 euros when compliance <80%, but €30 when compliance >80%). This script allows for automatically calculating either of this options.

To enable remuneration:

 1. Set `use_remuneration = TRUE`

 2. Choose a payment mode (`per_beep` or `threshold`)

 3. Adjust the corresponding payment values to your design
 
```{r set-remuneration}
payment_mode <- "per_beep"   # "per_beep" or "threshold"

payment_per_beep <- 0.50     # Used when payment_mode = "per_beep"

# Threshold payment structure:
# Names = compliance thresholds (0–1), values = payment amounts
payment_thresholds <- c(
  "0.80" = 30,   # If compliance >= 80% → €30
  "0.00" = 20    # Otherwise → €20
)
 
```


# 2. Folder Setup 

This section:

 - Creates required folders
 - Moves old files to /old
 - Checks that required Avicenna columns exist 
 - Warns about missing or invalid data

```{r setup-folders}

knitr::opts_chunk$set(echo = FALSE)

# Folder setup

required_folders <- c("data", "old", "exports")

for (f in required_folders) {
  if (!dir.exists(f)) {
    dir.create(f)
    message(paste("Created folder:", f))
  }
}

# Helper: Get newest file

get_latest_file <- function(path, pattern = "*.csv") {
  files <- list.files(path, pattern = pattern, full.names = TRUE)
  if (length(files) == 0) stop("No CSV files found in /data.")
  newest <- files[which.max(file.info(files)$mtime)]
  
  # Move older files to /old
  old_files <- setdiff(files, newest)
  if (length(old_files) > 0) {
    file.rename(old_files, file.path("old", basename(old_files)))
    message("Archived old data files to /old.")
  }
  
  return(newest)
}

```

# 3. Load data

If you exported **one Avicenna file**, the script will automatically load the newest .csv file in the /data folder.

If you exported **multiple Avicenna files** (i.e., different Activities), set use_multiple_files = TRUE and enter the Activity IDs in the User Inputs section in the first chunk.

If you set use_demo_dataset = TRUE, the code below will create some simulated data for you.


```{r load-data, message=FALSE}
library(tidyverse)
library(lubridate)


# ---- Incompatible settings check ----
if (use_demo_dataset && use_multiple_files) {
  cat("<div style='padding:12px; background:#F8D7DA; border-left:6px solid #C82333; margin:20px 0;'>
<strong>❌ Incompatible settings:</strong> <code>use_demo_dataset</code> and <code>use_multiple_files</code> are both set to <code>TRUE</code>.<br>
The demo dataset is a single simulated file and cannot be combined with multiple Avicenna activities.<br>
Please set one of them to <code>FALSE</code> and re-knit.
</div>")
  stop("Incompatible settings: use_demo_dataset and use_multiple_files cannot both be TRUE.")
}

if (use_demo_dataset && (demo_n_days != study_day || demo_beeps_per_day != beeps_per_day)) {
  cat("<div style='padding:12px; background:#FFF3CD; border-left:6px solid #FFA000; margin:20px 0;'>
<strong>⚠ Warning:</strong> Your demo settings (<code>demo_n_days</code>, <code>demo_beeps_per_day</code>) do not match your study settings (<code>study_day</code>, <code>beeps_per_day</code>).<br>
This will cause <code>n_completed</code> to exceed <code>expected</code> in the compliance table.<br>
Either align the demo settings with your study settings, or set <code>demo_n_days <- study_day</code> and <code>demo_beeps_per_day <- beeps_per_day</code>.
</div>")
}

# Load dataset (demo or real)
if (use_demo_dataset) {
  message("Using demo dataset with realistic compliance patterns...")

  set.seed(123)
  ids <- rep(1:demo_n_participants, each = demo_n_days * demo_beeps_per_day)
  n   <- length(ids)

  start_time      <- as.POSIXct(paste(start_date, "08:00:00"))
  scheduled_times <- as.POSIXct(unlist(lapply(1:demo_n_participants, function(p) {
    set.seed(p)
    as.numeric(start_time) + sort(sample(
      seq(0, demo_n_days * 24 * 3600, by = 3600),
      size    = demo_n_days * demo_beeps_per_day,
      replace = FALSE
    ))
  })), origin = "1970-01-01")

  status_values <- sample(
    x       = c(1, 0, 3, 2, 4, 6, 5),
    size    = n,
    replace = TRUE,
    prob    = c(0.70, 0.20, 0.05, 0.03, 0.01, 0.005, 0.005)
  )

  record_times <- ifelse(
    status_values %in% c(1, 6),
    as.numeric(scheduled_times) + runif(n, 1, 20) * 60,
    NA
  )

  esm <- tibble(
    Participant.ID          = ids,
    Session.Scheduled.Time  = scheduled_times,
    Prompt.Time             = scheduled_times,
    Record.Time             = as.POSIXct(record_times, origin = "1970-01-01"),
    Status                  = status_values
  )

  message("Demo dataset created.")

} else if (!use_multiple_files) {

  file <- get_latest_file(data_path)
  message(paste("Loading file:", file))
  esm <- read.csv(file) %>%
    mutate(
      Session.Scheduled.Time = ymd_hms(Session.Scheduled.Time),
      Prompt.Time            = ymd_hms(Prompt.Time),
      Record.Time            = ymd_hms(Record.Time),
      date                   = as.Date(Session.Scheduled.Time)
    )
  message("Loaded 1 Avicenna dataset.")

} else {

  if (length(activity_ids) == 0) {
    stop("use_multiple_files = TRUE but no activity_ids provided.")
  }

  message("Multiple-file mode: loading all activities listed in activity_ids...")

  get_latest_activity_file <- function(activity_id) {
    files <- list.files(data_path, pattern = as.character(activity_id), full.names = TRUE)
    if (length(files) == 0) stop(paste("No file found for Activity ID:", activity_id))
    files[which.max(file.info(files)$mtime)]
  }

  clean_activity <- function(df, activity_id) {
    df %>%
      mutate(
        Session.Scheduled.Time = ymd_hms(Session.Scheduled.Time),
        date                   = as.Date(Session.Scheduled.Time),
        activity_id            = activity_id
      ) %>%
      filter(date >= start_date) %>%
      select(Participant.ID, Session.Scheduled.Time,
             Prompt.Time, Record.Time, Status,
             activity_id, date)
  }

  activity_list <- lapply(activity_ids, function(id) {
    clean_activity(read.csv(get_latest_activity_file(id)), id)
  })

  esm <- bind_rows(activity_list)
  message(paste("Merged", length(activity_ids), "Avicenna activity files."))
}


```



# 4. Sanity checks

This section checks whether your dataset contains the required Avicenna variables and whether timestamps and dates look correct.
If something is wrong, the script will warn you and tell you what to fix.

**About the warnings below:**  
If something in your dataset looks unusual, this section will show a coloured message box.  
These warnings do *not* stop the document from knitting — they simply highlight issues that may affect your compliance calculations (for example: missing timestamps, unexpected Status values, or dates that don’t match your study setup).  

The coloured boxes are meant to be highly visible, so you can quickly spot potential problems without having to read through R warnings or code output.


Also, **a note on time zones**: Avicenna stores all timestamps internally in **UTC**. When exporting to CSV, Avicenna converts timestamps to local time — but depending on the device, DST rules, or the moment of export, this conversion may differ slightly from what you see in the dashboard.

This can lead to:

 - Some participants appearing to have prompts 1 hour earlier or later
 - Slight differences between the dashboard and the CSV
 - Apparent inconsistencies between participants

This is normal and does not affect the actual scheduling logic.
For more details, see Avicenna’s explanation:
https://forum.avicennaresearch.com/t/participation-period/1279#p-2906-time-zone-12

```{r sanity-checks, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
# Sanity checks (with human-readable messages)

required_cols <- c(
  "Participant.ID", "Session.Scheduled.Time",
  "Prompt.Time", "Record.Time", "Status"
)

missing <- setdiff(required_cols, names(esm))
if (length(missing) > 0) {
  cat("<div style='padding:10px; background:#F8D7DA; border-left:5px solid #C82333; margin:10px 0;'>
      <strong>❌ Error:</strong> Missing required Avicenna columns: ",
      paste(missing, collapse = ", "),
      "</div>
      
      ")
}

# Convert timestamps safely
suppressWarnings({
  esm <- esm %>%
    mutate(
      Session.Scheduled.Time = ymd_hms(Session.Scheduled.Time),
      Prompt.Time            = ymd_hms(Prompt.Time),
      Record.Time            = ymd_hms(Record.Time),
      date                   = as.Date(Session.Scheduled.Time)
    )
})

# Check for invalid Status values
valid_statuses <- c(0, 1, 2, 3, 4, 5, 6)
if (any(!esm$Status %in% valid_statuses)) {
  cat("<div style='padding:10px; background:#F8D7DA; border-left:5px solid #C82333; margin:10px 0;'>
      <strong>❌ Error:</strong> Unexpected Status values detected. Avicenna Status must be 0–6.
      </div>
      
      ")
}

# Check for impossible timestamps
if (any(esm$Record.Time < esm$Prompt.Time, na.rm = TRUE)) {
  cat("<div style='padding:10px; background:#FFF3CD; border-left:5px solid #FFA000; margin:10px 0;'>
      <strong>⚠ Warning:</strong> Some Record.Time values occur before Prompt.Time. This may indicate unsynced or corrupted entries.
      </div>
      
      ")
}

# Check for duplicate rows (skipped for demo data, where duplicates are expected)
has_duplicates <- !use_demo_dataset && any(duplicated(esm))
if (has_duplicates) {
  cat("<div style='padding:10px; background:#FFF3CD; border-left:5px solid #FFA000; margin:10px 0;'>
      <strong>⚠ Warning:</strong> Duplicate rows detected in the dataset. Consider removing them.
      </div>
      
      ")
}

# Check date range
if (min(esm$date, na.rm = TRUE) > start_date) {
  cat("<div style='padding:10px; background:#FFF3CD; border-left:5px solid #FFA000; margin:10px 0;'>
      <strong>⚠ Warning:</strong> Earliest date in data is AFTER the study start date.
      </div>
      
      ")
}

# Check for all-missing timestamps
if (all(is.na(esm$date))) {
  cat("<div style='padding:10px; background:#F8D7DA; border-left:5px solid #C82333; margin:10px 0;'>
      <strong>❌ Error:</strong> All timestamps are missing. Check your dataset.
      </div>
      
      ")
}

# If no issues at all:
if (length(missing) == 0 &&
    all(esm$Status %in% valid_statuses) &&
    !any(esm$Record.Time < esm$Prompt.Time, na.rm = TRUE) &&
    !has_duplicates &&
    !(min(esm$date, na.rm = TRUE) > start_date) &&
    !all(is.na(esm$date))) {

  cat("
<div style='padding:10px; background:#D4EDDA; border-left:5px solid #28A745; margin:20px 0;'>
<strong>✔ All checks passed.</strong> No issues detected in your dataset.
</div>
")
}
```


# 5. Compliance calculation

## A brief note on understanding compliance in Avicenna data

Each ESM prompt in Avicenna receives a Status code. These codes tell us what happened to each beep:

 - Unanswered/Unknown (ID 0) indicates that a response to this particular session hasn't been registered or uploaded yet. This status doesn't necessarily mean the session hasn't been initiated; it might have been prompted, but the response is yet to be recorded.
 - Completed (ID 1) indicates the participant completed the session and submitted it.
 - Canceled (ID 2) indicates the participant did not want to complete the session and explicitly canceled it.
 - Expired (ID 3) indicates the session was not responded to in time, so Avicenna automatically marked the session as expired.
 - Blocked (ID 4) indicates the session was blocked by another active session from the same Activity and could not be prompted.
 - Other (ID 5) indicates something else going on 
 - In Progress (ID 6) indicates the session is started by the participant but it's not completed, canceled, or expired yet.
 
Whilst running an ESM study in Avicenna, you may notice a lot of "0" (unknown/unanswered) responses. This does not mean that these responses will never come - it simply reflects the fact that the Avicenna server needs to connect to participants' phones to retrieve their data. You can expedite this process by asking participants to reload their app, or by manually reloading participants yourself, in the Participation tab in the Avicenna dashboard.

Now, in this template, we calculate **three different indicators** to help you understand where your study is at:

 - **Completed compliance**: This is a strict definition: Only `Status = 1` counts as a completed beep. Everything else counts as "not completed".
 - **Full status breakdown**: You will also get a  breakdown of these different status categories. This helps you understand why compliance may be low (is it because people really did not answer your prompts, or because their responses have just not yet come through?)
 - **Synchronization completeness**: the variable `fully_synced` tells you, for each participant, whether they are fully synced. A participant will get the value `1` for this variable if **every ESM prompt** has come through. If there is at least one prompt that's still in the air somewhere, a participant will get a value `0` for this variable.

```{r compliance-calc}
# Create status categories
esm <- esm %>%
  mutate(
    status_category = case_when(
      Status == 1 ~ "completed",
      Status == 0 ~ "unanswered",
      Status == 2 ~ "canceled",
      Status == 3 ~ "expired",
      Status == 4 ~ "blocked",
      Status == 5 ~ "other",
      Status == 6 ~ "in_progress",
      TRUE        ~ "other"
    ),
    is_completed = ifelse(Status == 1, 1, 0),
    response_delay_minutes = as.numeric(difftime(Record.Time, Prompt.Time, units = "mins"))
  )

# Expected number of beeps up to the current moment
expected_beeps <- ((study_day - 1) * beeps_per_day) + current_prompt

# Per-participant compliance
compliance_person <- esm %>%
  group_by(Participant.ID) %>%
  summarise(
    total_beeps = n(),
    n_completed = sum(is_completed),
    expected = expected_beeps,
    compliance = pmin(n_completed / expected, 1),   # strict compliance
    non_synced = sum(Status == 0),
    fully_synced = ifelse(non_synced == 0, 1, 0),
    .groups = "drop"
  )

# Add status breakdown per participant
status_breakdown <- esm %>%
  group_by(Participant.ID, status_category) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = status_category,
    values_from = n,
    values_fill = 0
  )

# Merge breakdown into compliance table
compliance_person <- compliance_person %>%
  left_join(status_breakdown, by = "Participant.ID")


```

The values below give you a quick overview of how compliant your sample is.
They are often the first numbers you want to check during a study.

```{r compliance-summary}
compliance_summary <- compliance_person %>%
  summarise(
    mean_compliance   = mean(compliance, na.rm = TRUE),
    median_compliance = median(compliance, na.rm = TRUE),
    sd_compliance     = sd(compliance, na.rm = TRUE),
    min_compliance    = min(compliance, na.rm = TRUE),
    max_compliance    = max(compliance, na.rm = TRUE)
  )

compliance_summary %>%
  knitr::kable(digits = 3, caption = "Compliance summary across all participants")

```

# 6. Plots

These visualizations help you quickly understand how your sample is responding.

They mostly reflect compliance, based on the description in the compliance calculation above, but also give some insight into the different statuses and mean response times.


```{r all-plots}

# Visualization Section

# Number of participants
n_participants <- length(unique(esm$Participant.ID))


# Unified Avicenna color palette
status_colors <- c(
  "unanswered"   = "#A6A6A6",
  "completed"    = "#4CAF50",
  "canceled"     = "#FFEB3B",
  "expired"      = "#F44336",
  "blocked"      = "#9C27B0",
  "in_progress"  = "#03A9F4",
  "other"        = "#B0BEC5"
)

# Helper: restrict palette to only the statuses present in the data
get_present_status_colors <- function(data, status_var = "status_category", palette = status_colors) {
  present <- intersect(unique(data[[status_var]]), names(palette))
  palette[present]
}

# PLOTS 

# 1. Status distribution (overall)
status_distribution_data <- esm %>% count(status_category)

status_distribution_plot <- ggplot(status_distribution_data,
                                   aes(x = status_category, y = n, fill = status_category)) +
  geom_col() +
  scale_fill_manual(values = get_present_status_colors(status_distribution_data)) +
  theme_minimal() +
  labs(title = "Overall Status Distribution", x = "Status", y = "Count")

print(status_distribution_plot)


# 2. Daily compliance over time (aggregated)
daily_mean_compliance <- esm %>%
  group_by(date) %>%
  summarise(compliance = mean(is_completed), .groups = "drop")

daily_mean_compliance_plot <- ggplot(daily_mean_compliance, aes(date, compliance)) +
  geom_line(color = "#4CAF50", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal() +
  labs(title = "Daily Mean Compliance", y = "Compliance")

print(daily_mean_compliance_plot)

# 3. Compliance distribution (histogram)
compliance_distribution_plot <- ggplot(compliance_person, aes(compliance)) +
  geom_histogram(bins = 20, fill = "#4CAF50", color = "white") +
  theme_minimal() +
  labs(title = "Compliance Distribution", x = "Compliance", y = "Participants")

print(compliance_distribution_plot)


# 4. Spaghetti plot (small vs large logic)
# 4. Spaghetti plot (small vs large logic)
daily_person <- esm %>%
  group_by(Participant.ID, date) %>%
  summarise(compliance = mean(is_completed), .groups = "drop")

n_participants <- length(unique(daily_person$Participant.ID))

if (n_participants == 0) {

  cat("<div style='padding:10px; background:#FFF3CD; border-left:5px solid #FFA000; margin:20px 0;'>
       <strong>⚠ No participant-level data available for the spaghetti plot.</strong>
       </div>")

} else if (n_participants <= 20) {

  palette_cols <- scales::hue_pal()(n_participants)
  names(palette_cols) <- unique(daily_person$Participant.ID)

  spaghetti_plot <- ggplot(daily_person, aes(date, compliance,
                                             group = Participant.ID,
                                             color = Participant.ID)) +
    geom_line(alpha = 0.8, linewidth = 1) +
    scale_color_manual(values = palette_cols) +
    scale_y_continuous(limits = c(0, 1)) +
    theme_minimal() +
    labs(title = "Daily Compliance per Participant", y = "Compliance", color = "Participant")

  print(spaghetti_plot)

} else {

  daily_person <- daily_person %>%
    mutate(compliance_bin = cut(
      compliance,
      breaks = c(-Inf, 0, 0.33, 0.67, 0.99, Inf),
      labels = c("0%", "1-33%", "34-66%", "67-99%", "100%"),
      right  = TRUE
    ))

  bin_colors <- c(
    "0%"     = "#F44336",
    "1-33%"  = "#FF7043",
    "34-66%" = "#FFC107",
    "67-99%" = "#8BC34A",
    "100%"   = "#4CAF50"
  )

  participant_order <- compliance_person %>%
    arrange(desc(compliance)) %>%
    pull(Participant.ID)

  daily_person$Participant.ID <- factor(
    daily_person$Participant.ID,
    levels = participant_order
  )

  spaghetti_plot <- ggplot(daily_person,
                           aes(x = date, y = Participant.ID, fill = compliance_bin)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_manual(values = bin_colors, name = "Daily compliance", drop = FALSE) +
    theme_minimal() +
    theme(
      axis.text.y  = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid   = element_blank()
    ) +
    labs(
      title    = "Daily Compliance per Participant",
      subtitle = paste0(n_participants, " participants, ordered by overall compliance (best at top)"),
      x        = "Date",
      y        = "Participant"
    )

  print(spaghetti_plot)
}


# 5. Response delay distribution (histogram)
response_delay_histogram <- esm %>%
  filter(!is.na(response_delay_minutes), response_delay_minutes >= 0) %>%
  ggplot(aes(response_delay_minutes)) +
  geom_histogram(bins = 30, fill = "#03A9F4", color = "white") +
  theme_minimal() +
  labs(title = "Response Delay Distribution", x = "Delay (minutes)", y = "Count")

print(response_delay_histogram)



# SMALL-SAMPLE PLOT (<50 participants)
if (n_participants < 50) {

  beep_distribution_status <- esm %>%
    group_by(Participant.ID, status_category) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(Participant.ID) %>%
    mutate(total_beeps = sum(n)) %>%
    ungroup()

  participant_order <- esm %>%
    group_by(Participant.ID) %>%
    summarise(prop_completed = mean(is_completed)) %>%
    arrange(desc(prop_completed)) %>%
    pull(Participant.ID)

  beep_distribution_status$Participant.ID <- factor(
    beep_distribution_status$Participant.ID,
    levels = participant_order
  )

  small_sample_status_plot <- ggplot(beep_distribution_status,
                                     aes(x = Participant.ID, y = n, fill = status_category)) +
    geom_col() +
    scale_fill_manual(values = get_present_status_colors(beep_distribution_status)) +
    theme_minimal() +
    labs(
      title = "Distribution of ESM Beeps per Participant by Status",
      x = "Participant ID",
      y = "Number of beeps",
      fill = "Status"
    ) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())

  print(small_sample_status_plot)
}



```



# 7. Optional Modules

These sections only run if you set the corresponding toggle to TRUE in the User Inputs section.

They allow you to:
 - Link external participant IDs
 - Generate additional visualizations
 - Export participant id's and compliance to Excel


## 7.1 External ID Matching

Sometimes the participant IDs used in Avicenna (the Participant.ID column) are not the same IDs you use in your own study administration — for example:

 - You assigned your own participant numbers
 - You use anonymized codes
 - You have a separate payment or contact list
 - You want to merge demographic or baseline questionnaire data

This optional module allows you to link Avicenna IDs to your own external IDs.

To use it:

 - Create a file named `external_ids.xlsx`
 - Place it in the same folder as this `.Rmd` file
 - Make sure it contains **at least** the following two columns:
 
#### Example structure of `external_ids.xlsx`:

```
Participant.ID,External.ID
145812,PP001
146889,PP002
147001,PP003
```

If you enable `use_external_ids = TRUE` in the first chunk, the script will automatically merge these IDs into your compliance table.

```{r id-matching, echo=FALSE, results='asis', message=FALSE, warning=FALSE}

if (use_external_ids) {

  library(readxl)

  # Read Excel file
  ids <- read_excel("external_ids.xlsx")

  # Check for required columns
  required_cols_ids <- c("Participant.ID", "External.ID")
  missing_cols <- setdiff(required_cols_ids, names(ids))
  if (length(missing_cols) > 0) {
    cat("
<div style='padding:12px; background:#F8D7DA; border-left:6px solid #C82333; margin:20px 0;'>
<strong>❌ Error:</strong> The external_ids.xlsx file is missing required columns:<br>
", paste(missing_cols, collapse = ", "), "<br>
Please add these columns and re-knit.
</div>

")
    stop("external_ids.xlsx is missing required columns: ", paste(missing_cols, collapse = ", "))
  }

  # Ensure matching types
  ids <- ids %>% mutate(Participant.ID = as.character(Participant.ID))
  compliance_person <- compliance_person %>% mutate(Participant.ID = as.character(Participant.ID))

  # Duplicate ID check
  if (any(duplicated(ids$Participant.ID))) {

    dups <- ids$Participant.ID[duplicated(ids$Participant.ID)]

    cat("
<div style='padding:12px; background:#FFF3CD; border-left:6px solid #FFA000; margin:20px 0;'>
<strong>⚠ Warning:</strong> Duplicate Participant.ID values found in external_ids.xlsx.<br>
The following IDs appear more than once:<br>
", paste(unique(dups), collapse = ", "), "<br>
Only the first occurrence for each participant will be used.
</div>

")

    ids <- ids %>% distinct(Participant.ID, .keep_all = TRUE)
  }

  # Merge
  compliance_person <- compliance_person %>%
    left_join(ids, by = "Participant.ID")
}
```





## 7.2 Excel Export 

This module creates a clean Excel file containing: Participant ID, total beeps, number of completed beeps (`n_completed`), compliance proportion, sync status, rank (highest to lowest compliance), and optional remuneration (if enabled).

```{r excel-export}

if (use_excel_export) {
  library(openxlsx)
  
    compliance_person <- compliance_person %>%
    mutate(Participant.ID = as.character(Participant.ID))

  # ---- 1. Prepare main export data ----
  base_cols <- c("Participant.ID", "n_completed", "expected", "compliance",
                 "non_synced", "fully_synced")

  if ("External.ID" %in% names(compliance_person)) {
    base_cols <- c("Participant.ID", "External.ID", base_cols[-1])
  }

  export_data <- compliance_person %>%
    select(all_of(base_cols)) %>%
    arrange(desc(compliance)) %>%
    mutate(
      compliance = pmin(compliance, 1),   # cap here so export is always clean
      rank = row_number()
    )

  # ---- 2. Prepare payment sheet (optional) ----
  if (use_remuneration) {

    if (payment_mode == "per_beep") {

      payment_data <- export_data %>%
        transmute(
          Participant.ID,
          External.ID = if ("External.ID" %in% names(.)) External.ID else NA_character_,
          compliance,
          remuneration = n_completed * payment_per_beep
        )

    } else if (payment_mode == "threshold") {

      thresholds <- as.numeric(names(payment_thresholds))

      payment_data <- export_data %>%
        rowwise() %>%
        mutate(
          remuneration = {
            eligible <- thresholds[round(compliance, 6) >= thresholds]
            if (length(eligible) == 0) {
              payment_thresholds[which.min(thresholds)]
            } else {
              payment_thresholds[as.character(max(eligible))]
            }
          }
        ) %>%
        ungroup() %>%
        transmute(
          Participant.ID,
          External.ID = if ("External.ID" %in% names(.)) External.ID else NA_character_,
          compliance,
          remuneration
        )
    }
  }

  # ---- 3. Create workbook ----
  wb <- createWorkbook()

  addWorksheet(wb, "Compliance")
  writeData(wb, "Compliance", export_data)

  if (use_remuneration) {
    addWorksheet(wb, "Payments")
    writeData(wb, "Payments", payment_data)
  }

  # ---- 4. Export directory ----
  export_dir <- file.path(getwd(), "exports")
  if (!dir.exists(export_dir)) dir.create(export_dir)

  timestamp_file <- paste0("ESM_Compliance_", format(Sys.time(), "%Y-%m-%d_%H-%M"), ".xlsx")
  latest_path    <- file.path(export_dir, "ESM_Compliance_latest.xlsx")

  if (file.exists(latest_path)) {
    archive_name <- paste0("ARCHIVED_", format(file.info(latest_path)$mtime, "%Y-%m-%d_%H-%M"), ".xlsx")
    file.rename(latest_path, file.path(export_dir, archive_name))
  }

  saveWorkbook(wb, file = file.path(export_dir, timestamp_file), overwrite = TRUE)
  saveWorkbook(wb, file = latest_path, overwrite = TRUE)

  message("Excel export created in the 'exports' folder.")
}

```

